Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-01-11 scagnostics.Rmd
More references added.
start of index computation section

The term "projection pursuit" was coined by @FT74 to describe a procedure for searching high (say $p-$)dimensional data for "interesting" low-dimensional projections ($d=1$ or $2$ usually). The procedure, originally suggested by @kr69, involves defining a criterion function, or index, that measures the "interestingness" of each $d$-dimensional projection of $p$-dimensional data. This criterion function is optimized over the space of all $d$-dimensional projections of $p$-space, searching for both global and local maxima. It is hoped that the resulting solutions reveal low-dimensional structure in the data not found by methods such as principal component analysis.

@@

# Constructing a projection pursuit index
\label{sec:construct}

A projection pursuit index (PPI) is a scalar function $f$ defined on an $d$-dimensional data distribution. Typically the definition is such that larger values of $f$ indicate a more interesting distribution of observations, and therefore maximizing $f$ over all possible $d$ dimensional projections of a data set with $p>d$ parameters will find the most interesting projections. As most indices characterize deviations from a normal distribution, one would first map the empiric data distribution onto a density for which such deviations can then be defined in different ways, see e.g. @CBC93. When departing from the idea of characterizing differences to the normal distribution it may however be preferred to work directly with the empirical data distribution and instead characterize it by measures based on e.g. the minimal spanning tree or the mutual information.

## Standardization
Lower dimensional projections are highly sensitive to standardization performed on the original distribution. We want to avoid e.g.  highlighting directions based only on different scales, or artificially extending the parameter range by including outliers, which may result in seemingly linear behavior for most points. Methods to consider include rescaling of individual directions to a common interval (e.g. to fall between $[0,1]$), sphering, outlier removal or transformation to a logarithmic scale. The method(s) of choice should be informed by the distributions found in the data set as well as aims of the analysis. \textcolor{red}{particular challenges: noise only directions, identification of outliers?}
\textcolor{red}{for holes/cmass index they are only well defined when data is standardized, how about the other index functions?}
\textcolor{red}{When keeping in mind the concrete example of posterior samples outliers should not be an issue, if extreme outliers are found one should probably investigate origin. Also noise only directions, e.g. nuisance parameters, are also not an issue as they contain no structure, see BBH example.}

\textcolor{red}{\cite{JS87} discusses centering and sphering of the data points. Centering is needed if index is not translational invariant (should not be the case for any index considered here). Arguments for sphering are that scale effects should be extracted, we don't want to reproduce information already obtained by PCA. To my mind rescaling to an overall interval is more appropriate here. Rotations between parameters should not affect the outcome, but will make interpretation more difficult. Standardizing the variance will remove information about the posterior variance. Overall we are interested in strong association only, if these structures were affected by the standardization procedure, it would most likely be unwelcome effects?}

>>>>

The term "projection pursuit" (PP) was coined by @FT74 to describe a procedure for searching high (say $p-$)dimensional data for "interesting" low-dimensional projections ($d=1$ or $2$ usually). The procedure, originally suggested by @kr69, involves defining a criterion function, or index, that measures the "interestingness" of each $d$-dimensional projection of $p$-dimensional data. This criterion function is optimized over the space of all $d$-dimensional projections of $p$-space, searching for both global and local maxima. It is hoped that the resulting solutions reveal low-dimensional structure in the data not found by methods such as principal component analysis. Projection pursuit is primarily used for visualization, with projected data always reported as plots.

@@

# Projection pursuit index construction and optimization
\label{sec:construct}

A projection pursuit index (PPI) is a scalar function $f$ defined on an $d$-dimensional data set, computed by taking a $d$-dimensional projection of an $n\times p$ data matrix. Typically the definition is such that larger values of $f$ indicate a more interesting distribution of observations, and therefore maximizing $f$ over all possible $d$ dimensional projections of a data set with $p>d$ variables will find the most interesting projections. This section describes the 12 indexes that are to be used to explore bivariate association. Some data pre-processing, including standardization, is advisable, prior to optimizing the PPI.

<!-- As most indices characterize deviations from a normal distribution, one would first map the empiric data distribution onto a density for which such deviations can then be defined in different ways, see e.g. @CBC93. When departing from the idea of characterizing differences to the normal distribution it may however be preferred to work directly with the empirical data distribution and instead characterize it by measures based on e.g. the minimal spanning tree or the mutual information.-->

## Scaling and standardization

Making a plot always involves some choice of scaling. When a scatterplot is made, effectively, albeit under the hood, the data is scaled into a range of $[a, b]$ (often $a=0, b=1$) on both axes to print it on a page or display in a graphics device window. The range deliminates page space within which to draw. The upshot is that the original data scale is standardized to the range and aspect ratio on the display space. It maybe that the original range of X is $[1, 1000]$ and Y is $[1, 1.6]$ but the display linearly warps this to $[0, 1]$ and $[0, 1]$, giving both variables equal visual weight.

With high dimensional data, and particularly projections, it is also necessary to re-scale the original range, and it is important to pay attention to what is conventional, or possible, and the effects. The PPIs also may require specific scaling for them to be effectively computed. Both of these are addressed here. The common pre-processing include:

- Standardizing each variable, to have mean 0 and variance 1, so that individual variable scales do not affect the result. Different variable scales are examinable without resorting to projection pursuit, so can be handled prior to searching through high dimensions.
- Sphering the high-dimensional data is often done to remove linear dependence. This is typically done using principal component analysis, and using the principal components as the variables passed to PP. If linear dependence is the only structure PP is not needed, and thus this is removed before PP so that the PPIs are not distracted by simple structure.
- Transform single variables to reduce skewness. It is marginal structure, visible in a single variable, which doesn't need a multivariate technique to reveal. Skewed distributions will inadvertently affect the PPIs, distracting the search for dependence. 
- Remove outliers, which may be an iterative process, to discover, identify and delete. Extreme values will likely affect PPI performance. Outliers can be examined on a case by case basis later.
- Possibly remove noise dimensions, which is also likely to be an iterative process. Directions where the distribution is purely noise make optimization of a PPI more difficult. If a variable is suspected to have little structure and relationship with other variables, conducting PP on the subset of variables without them may improve the efficiency of the search.
- Centering and scaling of the projected data, can be helpful visually. If the data has a small amount of non-normal distribution in some directions, the projected data can appear to wander around the plot window during a tour. It doesn't matter what the center of the projected data is, so centering removes a wandering scatterplot. Less commonly, it may be useful to scale the projected data to standard values, which would be done to remove any linear dependence remaining in the data.

<!--
Lower dimensional projections are highly sensitive to standardization performed on the original distribution. We want to avoid e.g.  highlighting directions based only on different scales, or artificially extending the parameter range by including outliers, which may result in seemingly linear behavior for most points. Methods to consider include rescaling of individual directions to a common interval (e.g. to fall between $[0,1]$), sphering, outlier removal or transformation to a logarithmic scale. The method(s) of choice should be informed by the distributions found in the data set as well as aims of the analysis. \textcolor{red}{particular challenges: noise only directions, identification of outliers?}
\textcolor{red}{for holes/cmass index they are only well defined when data is standardized, how about the other index functions?}
\textcolor{red}{When keeping in mind the concrete example of posterior samples outliers should not be an issue, if extreme outliers are found one should probably investigate origin. Also noise only directions, e.g. nuisance parameters, are also not an issue as they contain no structure, see BBH example.}

\textcolor{red}{\cite{JS87} discusses centering and sphering of the data points. Centering is needed if index is not translational invariant (should not be the case for any index considered here). Arguments for sphering are that scale effects should be extracted, we don't want to reproduce information already obtained by PCA. To my mind rescaling to an overall interval is more appropriate here. Rotations between parameters should not affect the outcome, but will make interpretation more difficult. Standardizing the variance will remove information about the posterior variance. Overall we are interested in strong association only, if these structures were affected by the standardization procedure, it would most likely be unwelcome effects?}
-->

## Notation
